Estimating duration in the presence of misclassification

### 2024-03-07 introduction.tex
proof intro

COVID-19 (coronavirus disease 2019), the disease caused by the virus SARS-CoV-2 (severe acute respiratory syndrome coronavirus 2), is estimated to have killed approximately 15 million in the acutest phase (to the end of 2021) and 30 million in total~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
While would be reassuring if this was a rare or extreme event, but history suggests otherwise (see \cref{intro:fig:pandemic-timeline}).
\begin{figure}
    \includegraphics[width=\textwidth]{introduction/pandemic-timeline}
    \caption[Timeline of pandemics.]{%
        Timeline of pandemics.
        See \url{https://ourworldindata.org/historical-pandemics} for details and sources.
        Licensed under \href{https://creativecommons.org/licenses/by/4.0/}{CC-BY} by Saloni Dattani, Klara Auerbach, Marwa Boukarim, and Max Roser.
    }
    \label{intro:fig:pandemic-timeline}
\end{figure}

Fully quantifying the burden of historical pandemics is challenging due to the absence of systematic records.
However, our best records suggests that, since 1600, they have (on average) occurred every four years with a mean death toll of 5.6 million~\autocite{maraniNovelEpidemics}.
Modern estimates suggest that we should expect over 2.5 million deaths per year from pandemics~\autocite{madhavPandemicMortality}.

@@

The \emph{effective reproduction number}, $\Re$, is the mean number of secondary infections generated by a single infection at time $t$.
In particular, it uses the behaviour and immunity status at time $t$.
With constant behaviour, $\Re$ declines smoothly as individuals in the population become immune due to infection.
$\Re$ is useful because it describes the current dynamics of the epidemic: $\Re > 1$ means the epidemic is growing, $\Re < 1$ means it is shrinking, and $\Re = 1$ means it is stable.
In the simplest epidemic models (see \cref{E-SEIR:sec:SIR}), $\Re$ is proportional to the product of the rate of contacts and the proportion of the population that is susceptible.
Therefore, it informs the strength of intervention (either reducing contacts or amount of immunisation) required to control the epidemic.

@@

Community testing is broadly available, but does not test a random sample of the population.
The number of tests conducted varies over time and space, for a variety of known and unknown reasons.
These include: policy on who is eligible for testing, the availability of tests, and the willingness of individuals to be tested.
Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
For large parts of the pandemic, community testing was widely available in the UK meaning that this is a large dataset, allowing detailed analyses to be performed, including stratification by small geographic areas or age groups.

Seroprevalence testing is a marker of recent prior infection.
Prior infection is more common than current infection, hence seroprevalence estimation requires a smaller sample size than current prevalence.
However, there is a delay between infection and the development of antibodies, hence seroprevalence is not a real-time measure of the burden of the disease.

Severe events are generally a reliable measure of the burden of the disease.
The criteria for being a severe event are well-defined and generally well-recorded in the UK context.
Therefore, the data is subject to fewer biases than community testing.
However, SARS-CoV-2 rarely leads to severe events in younger individuals, meaning severe events only capture transmission to a subset of the population.

@@

A major advantage of the CIS over REACT is that the CIS has daily testing of individuals, while REACT has no test results between rounds.
This means that REACT cannot be used to estimate transmission or incidence between rounds, instead relying on model assumptions at these times.
REACT missed the peak in late December 2020, which complicated analysis of the survey data at that time~\autocite{rileyREACTround8}.

@@

Furthermore, prevalence surveys are noisy because they are based on population sampling.
This noise needs to be considered in the application of backcalculation (see \cref{E-inc-prev:sec:observation-process}).

The original formulations of backcalculation neglected the observation process.
AIDS is almost always diagnosed and recorded in developed countries.
That is, the observations of AIDS diagnoses are the same as the true incidence of AIDS.
An exception is that the reporting delay may be important~\autocite{paganoHIV}; but reporting delay is not important for this thesis's work.
Since its development, backcalculation has been further developed for a wide-variety of contexts.
% Reporting delay is not relevant to the studies in this thesis, and I ignore it in this and further discussion.

@@

Compartmental models split the population into compartments, each representing a different state of the disease.
The \emph{dynamics} of the epidemic, how it changes over time, are described by the transitions between compartments.
Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
Unlike renewal equations, there are not widely-used implementations of compartmental models.
Most of the implementations used by SPI-M-O members were in the SEIR (susceptible, exposed, infectious, recovered) framework.
The SEIR framework, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components.
The SEIR framework is described in detail in \cref{E-SEIR:sec:mechanistic-models}.
Compartmental models can be fit to almost any data source, and commonly incorporate multiple data sources, known as \emph{data synthesis}, a technique reviewed by \textcite{birrellEvidence}.
Most compartmental models used by SPI-M-O were age-stratified, as differences between age groups are important for both the transmission and severity of SARS-CoV-2 (see \cref{E-SEIR:sec:structured-populations}).

@@

Each week, the members of SPI-M would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
The nowcasting estimates were based on mechanistic models.
These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value~\autocite{parkCombining}.
The consensus value was published on \url{gov.uk} and informed the government's response to the epidemic~\autocite{govRnumber}.
The academic groups contributing estimates used a variety of models, reviewed by \textcite{royalSocietyRnumber}; see appendix 2 of that review for details of the specific implementations.

@@

Two approaches are commonly used to estimate the duration of a disease: either modelling an underlying biomarker or modelling the duration directly~\autocite{sweetingEstimating}.
For SARS-CoV-2, both approaches would be based on the results of RT-PCR testing (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}).

Modelling a biomarker, as I do in \cref{E-ATACCC}, models how the amount of virus in an individual varies over time.
The RT-PCR tests commonly used give a proxy for viral load for positive tests (see \cref{E-biology-data:sec:PCR-process}).
Modelling a biomarker uses more information per observation, but requires stronger assumptions.
For example, a functional form of the behaviour over time of the biomarker is required.
In addition, if individual-level variation wants to be included, which is generally important, a distributional form of random effects is required.
This requires both sufficient observations per individual and an assumption on the form of the distribution.

@@

An early meta-analysis estimated that individuals test positive for SARS-CoV-2 for 14.6 days (95\% CI: 9.3--20.0 days)~\autocite{cevikShedding}.
The maximum observed duration was 83 days, although beyond day 9 no live virus was detected, indicating that the individuals testing positive for longer than this were not infectious.
However, the majority of the studies included in the meta-analysis involved hospitalised patients or had unclear inclusion criteria, and hence may not be representative of the general population.
These summary estimates are insufficient for backcalculation, which requires a distribution of the duration of positivity.

@@

For the purposes of this thesis, estimates applicable to the general population are required.
Therefore, I estimate the duration of positivity from the ATACCC and CIS studies, as described above.

@@

All inference in this thesis is Bayesian.
Bayesian inference has several relevant advantages.
Prior information allows identifiability in situations where this would not otherwise be possible, allows for combining knowledge from multiple sources, and many previous studies I build used Bayesian inference.

@@

The overall question that I investigate in this thesis is: how can prevalence surveys best be used to understand incidence and transmission?
Prevalence surveys provide data that is far more representative of the general population than other data sources (see \cref{intro:sec:data-sources}), therefore, conclusions based on them should be far more robust.
There are two components to this question.
First, how can prevalence surveys be designed to help estimate these quantities.
Second, what statistical methodology can be used to estimate these quantities from prevalence surveys.
This thesis will focus on the second component.
However, natural extensions of the work I undertake, will inform the first component.

@@

Next, in \cref{E-perf-test}, I estimate the duration of RT-PCR-positivity from double censored data with arbitrary periods where infections could be missed (a generalisation of truncation).
This is based on the CIS data.
While statistical frameworks have been previously developed which included this case, they had only been applied in special cases.
In \cref{E-imperf-test}, I extend this framework to include false negatives.

@@

Finally, in \cref{E-SEIR}, I estimate the incidence of SARS-CoV-2 in England from only the CIS data.
This includes the development of a novel observation model to link the mechanistic transmission model to the data.
The observation model allows the inclusion of the long tail in the duration of positivity, without being computationally infeasible.

@@

In \cref{E-biology-data}, I provide background information on the natural history of SARS-CoV-2, the RT-PCR testing process used to produce the data used throughout this thesis, and describe the two cohort studies used in this thesis.
Then, \cref{E-inc-prev}, I formally describe the stochastic processes which generate infections, prevalence, and the observations of prevalence.
A key insight in this chapter is that the distribution of the duration of positivity is the key quantity linking incidence and prevalence.

Therefore, I then turn to estimating this distribution.
I start, in \cref{E-ATACCC}, by estimating the duration of positivity from the ATACCC study.
This provides a reliable estimate of the bulk of the distribution (out to around 20 days) but shows that there is a significant tail.
To estimate the tail, a large sample size and long follow-up is required.
Therefore, I turn to the CIS data in \cref{E-perf-test,E-imperf-test}.
In \cref{E-perf-test} I provide a framework for estimating the duration of positivity from the CIS data, assuming perfect tests.
However, the assumption of perfect tests is not realistic; in \cref{E-imperf-test} I extend the framework to include false negatives.

>>>>

COVID-19 (coronavirus disease 2019), the disease caused by the virus SARS-CoV-2 (severe acute respiratory syndrome coronavirus 2), is estimated to have killed approximately 15 million in the acutest phase (to the end of 2021) and 30 million in total~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
History suggests severe pandemics are not rare (see \cref{intro:fig:pandemic-timeline}).
\begin{figure}
    \includegraphics[width=\textwidth]{introduction/pandemic-timeline}
    \caption[Timeline of pandemics.]{%
        Timeline of pandemics.
        See \url{https://ourworldindata.org/historical-pandemics} for details and sources.
        Licensed under \href{https://creativecommons.org/licenses/by/4.0/}{CC-BY} by Saloni Dattani, Klara Auerbach, Marwa Boukarim, and Max Roser.
    }
    \label{intro:fig:pandemic-timeline}
\end{figure}

Fully quantifying the burden of historical pandemics is challenging due to the absence of systematic records.
However, our best records suggests that, since 1600, they have (on average) occurred every four years with a mean death toll of 5.6 million~\autocite{maraniNovelEpidemics}.
Recent estimates suggest that the current mean impact of pandemics is over 2.5 million deaths per year~\autocite{madhavPandemicMortality}.

@@

The \emph{effective reproduction number}, $\Re$, is the mean number of secondary infections generated by a single infection at time $t$.
In particular, it uses the behaviour and immunity status at time $t$.
With constant behaviour and a disease that grants immunity following infection, $\Re$ declines smoothly.
$\Re$ is useful because it describes the current dynamics of the epidemic: $\Re > 1$ means the epidemic is growing, $\Re < 1$ means it is shrinking, and $\Re = 1$ means it is stable.
In simple epidemic models (see \cref{E-SEIR:sec:SIR}), $\Re$ is proportional to the product of the rate of contacts, the riskiness of those contacts, and the proportion of the population that is susceptible.
Therefore, it informs the strength of intervention, \eg social distancing or vaccination, required to control the epidemic.

@@

Community testing is broadly available, but does not test a random sample of the population.
The number of tests conducted varies over time and space, for a variety of known and unknown reasons.
These include: policy on who is eligible for testing, the availability of tests, and the willingness of individuals to be tested.
Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
For large parts of the pandemic, community testing was widely available in the UK.
Therefore, this is a large dataset, allowing detailed analyses to be performed, including stratification by small geographic areas or age groups.

Seroprevalence testing is a marker of recent prior infection.
Prior infection is more common than current infection, hence seroprevalence estimation requires a smaller sample size than current prevalence.
However, there is a delay between infection and the development of antibodies, hence seroprevalence is not a real-time measure of the burden of the disease.
Furthermore, seroprevalence data is normally based on convenience samples (\eg blood donors) who are not representative of the general population.

Severe events are generally a reliable measure of the burden of the disease.
The criteria for being a severe event are normally well-defined and well-recorded in the UK context.
Therefore, the data is subject to fewer biases than community testing.
However, SARS-CoV-2 rarely leads to severe events in younger individuals, meaning severe events only capture transmission to a subset of the population.

@@

A major advantage of the CIS over REACT is that the CIS has daily testing of individuals, while REACT has no test results between rounds.
This means that REACT cannot be used to estimate transmission or incidence between rounds, instead relying on model assumptions at these times.
A particular instance of this is that a peak in late December 2020 occurred between rounds of the REACT survey, complicating analysis of the survey data at that time~\autocite{rileyREACTround8}.

@@

Furthermore, prevalence surveys are noisy because they are based on population sampling.
This observation noise needs to be considered in the application of backcalculation (see \cref{E-inc-prev:sec:observation-process}).

The original formulations of backcalculation neglected any observation noise.
AIDS is almost always diagnosed and recorded in developed countries.
That is, the observations of AIDS diagnoses are the same as the true incidence of AIDS.
An exception is that the reporting delay may be important~\autocite{paganoHIV}; but reporting delay is not important for this thesis's work.
Since its development, backcalculation has been further developed for a wide-variety of contexts.
% Reporting delay is not relevant to the studies in this thesis, and I ignore it in this and further discussion.

@@

Compartmental models split the population into compartments, each representing a different state of the disease.
The \emph{dynamics} of the epidemic, how it changes over time, are represented by the transitions between compartments.
Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
Unlike renewal equations, there are not widely-used implementations of compartmental models.
A commonly used frameork is the SEIR (susceptible, exposed, infectious, recovered) model.
The SEIR model, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components; it is described in detail in \cref{E-SEIR:sec:mechanistic-models}.

Compartmental models can be fit to almost any data source, and commonly incorporate multiple data sources, known as \emph{data synthesis}, a technique reviewed by \textcite{birrellEvidence}.

@@

Each week, the members of SPI-M would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
The nowcasting estimates were based on mechanistic models.
These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value~\autocite{parkCombining}.
The consensus value was published on \url{gov.uk} and informed the government's response to the epidemic~\autocite{govRnumber}.
The academic groups contributing estimates used a variety of models, reviewed by \textcite{royalSocietyRnumber}; see appendix 2 of that review for details of the specific implementations.
Most compartmental models used by SPI-M-O were age-stratified, as differences between age groups are important for both the transmission and severity of SARS-CoV-2 (see \cref{E-SEIR:sec:structured-populations}).

@@

Two approaches are commonly used to estimate the duration of a disease: either modelling an underlying biomarker or modelling the duration directly~\autocite{sweetingEstimating}.
For SARS-CoV-2, both approaches would be based on the results of RT-PCR testing (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}\todo{move to where RT-PCR / PCR is first mentioned}).

The biomarker approach, which I take in \cref{E-ATACCC}, models how the amount of virus in an individual varies over time.
The RT-PCR tests commonly used give a proxy for viral load for positive tests (see \cref{E-biology-data:sec:PCR-process}).
Modelling a biomarker uses more information per observation, but requires stronger assumptions.
For example, a functional form of the behaviour over time of the biomarker is required.
In addition, if individual-level variation wants to be included, which is generally important, a distributional form of random effects is required.
This requires both sufficient observations per individual and an assumption on the form of the distribution.

@@

An early meta-analysis estimated that individuals test positive for SARS-CoV-2 for a mean of 14.6 days (95\% CI: 9.3--20.0 days)~\autocite{cevikShedding}.
The maximum observed duration was 83 days, although beyond day 9 no live virus was detected, indicating that the individuals testing positive for longer than this were not infectious.
However, the majority of the studies included in the meta-analysis involved hospitalised patients or had unclear inclusion criteria, and hence may not be representative of the general population.
These summary estimates are insufficient for backcalculation, which requires a distribution of the duration of positivity.

@@

For the purposes of this thesis, estimates applicable to the general population are required.
There are no reliable estimates of the full distribution of the duration of positivity in a general population cohort available in the literature.
Therefore, I estimate the duration of positivity from the ATACCC and CIS studies, as described above.

@@

All inference in this thesis is Bayesian.
Including prior information allows identifiability in situations where this would not otherwise be possible and allows for combining knowledge from multiple sources.
Many previous studies I build on used Bayesian inference.

@@

The overall question that I investigate in this thesis is: how can prevalence surveys best be used to understand incidence and transmission?
Prevalence surveys provide data that is far more representative of the general population than other data sources (see \cref{intro:sec:data-sources}), therefore, conclusions based on them should be far more robust.
There are two components to this question.
First, how can prevalence surveys be designed to help estimate these quantities.
Second, what statistical methodology can be used to estimate these quantities from prevalence surveys.
This thesis will focus on the second component.
However, natural extensions of the work I undertake would inform the first component.

@@

Next, in \cref{E-perf-test}, I estimate the duration of RT-PCR-positivity from double censored data with arbitrary periods where infections could be missed (a generalisation of truncation).
This is based on the CIS data.
While statistical frameworks have been previously developed which included this case, they had not been used in practice.
In \cref{E-imperf-test}, I extend this framework to include false negatives.

@@

\todo[inline]{Check what methodological claims I want to make about ch 7}
Finally, in \cref{E-SEIR}, I estimate the incidence of SARS-CoV-2 in England from only the CIS data.
This includes the development of a novel observation model to link the mechanistic transmission model to the data.
The observation model allows the inclusion of the long tail in the duration of positivity, without being computationally infeasible.

@@

In \cref{E-biology-data}, I provide background information on the natural history of SARS-CoV-2, the RT-PCR testing process used to produce the data used throughout this thesis, and describe the two cohort studies used in this thesis.
Then, in \cref{E-inc-prev}, I formally describe the stochastic processes which generate infections, prevalence, and the observations of prevalence.
A key insight in this chapter is that the distribution of the duration of positivity is the key quantity linking incidence and prevalence.

Therefore, I then turn to estimating this distribution.
I start, in \cref{E-ATACCC}, by estimating the duration of positivity from the ATACCC study.
This provides a reliable estimate of the bulk of the distribution (out to around 20 days) but shows that there is a significant tail.
To estimate the tail, a large sample size and long follow-up is required.
Therefore, I turn to the CIS data in \cref{E-perf-test,E-imperf-test}.
In \cref{E-perf-test}, I provide a framework for estimating the duration of positivity from the CIS data, assuming perfect tests.
However, the assumption of perfect tests is not realistic; in \cref{E-imperf-test}, I extend the framework to include false negatives.
